Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-11-19 paper/main.tex
editted history and statistics

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5 stream, in the last 5~Gyr, which -- we recall the reader -- is the time interval investigated in our simulations. The results of this analysis are reported in Fig.~\ref{fig:histogram_impact_time} and in Table~xx \paola{I suggest to report the results of Fig 3 also in a Table, in which we report for each GC which produced a gap, the time of close passage and the total number of gaps it generated.} In total, we report the finding of 73 gaps in our 50 simulations, that is 1.5 gaps/simulation, on average, created by 18 different perturbers. NGC~2808 impacts Palomar 5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition}  in each case. Since this interaction occurred less than one orbital period ago, the orbital solutions remain similar, producing consistent results across all simulations, within the uncertainties.  For encounters which took place at earlier times, the clusters responsible of creating the gaps may vary -- that is, gaps are not created at the same time in all simulations and the parameters of a specific encounter between a  globular cluster and Palomar~5 stream may vary significantly from a simulation to another with the turning back of time, so that in some cases a cluster may produce a gap and in some others the perturbation is not strong or close enough to produce any detectable signature in the Palomar~5 stream.  In total, beside NGC~2808, we report the creation of 24 gaps in Pal~5 tails, over the last 3~Gyr generated by \paola{15: check this value, I derived it by eye} different clusters, which corresponds to a rate of 0.8 gaps/100~Myr (note that, including NGC~2808, this rate increases to 29.5 gaps/100~Myr). At even earlier times, the probability of creating gaps in Palomar~5 streams diminishes significantly, but one must take into account that at these times the tails of Pal~5 have not yet had time to develop sufficiently, and the probability of close encounters with other clusters consequently decreases. Overall, across 50 simulations, as previously stated, we observe 73 gaps. The distribution of simulations with a given number of gaps is provided in Table~\ref{table:gap_distribution}.  If we consider NGC~2808 to be an outlier and exclude it from the experiment, we observe an average of 0.6 gaps per simulation. Including all observed gaps results in an average of 1.5 gaps per simulation. Interestingly, most gaps appear in the leading tail, which we consider coincidental \paola{I am still surprised by the high occurrence of gaps in the leading tail, but I amm still not able to come up with an explanation for this. I mean, is it really coincidental? I propose to leave the discussion as it is for the moment. We will add more, if we come up with an other possible explanation}, as stars escape the cluster's L1 and L2 points at similar rates, and the small distance difference between tails' galactocentric distance is negligible. We expect gaps to be equally likely in both tails. However, only 8 of the 73 gaps occur in the trailing tail, with 44 linked to NGC~2808's flyby. After removing these, there are 21 gaps in the leading tail and 8 in the trailing. By the binomial distribution, the probability of observing up to 8 successes in 29 trials, given a 50\% chance of success, is 1.2\%—unlikely, but possible. Additionally, other perturbers impact at consistent times, which may also violate the assumption of independent events, as seen with NGC~2808.

@@

    Following  \citet{2012ApJ...748...20C} we can also estimate the rate at which gaps are created per unit stream length. For our simulations, this rate is given by: \begin{equation} \label{eq:gap_creation_rate} \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the Dirac delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}

    This computation allows us to analyze the distribution of gap creation rates across all simulations. The results are shown in Fig.~\ref{fig:gapcreationrate}. This is roughly consistent with a simple estimate of the average gap creation rate: with 72\paola{73?} gaps over 5 Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.015~km~s$^{-1}$~kpc$^{-2} = 0.015~\rm{Gyr^{-1}kpc^{-1}}$ . This is about double the weighted mean gap creation rate of 0.009~km~s$^{-1}$~kpc$^{-2}$. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}, where the likelihood of impacting a shorter stream is lower.

@@

>>>>

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5's stream, in the last 5~Gyr, which -- we recall the reader -- is the time interval investigated in our simulations. The results of this analysis are reported in Fig.~\ref{fig:histogram_impact_time} and in Table~xx \paola{I suggest to report the results of Fig 3 also in a Table, in which we report for each GC which produced a gap, the time of close passage and the total number of gaps it generated.}. NGC~2808 impacts Palomar 5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition} in each case. Since this interaction occurred less than one orbital period ago, the uncertainties lead to orbital solutions remain that similar and thus producing consistent results across all simulations. However, as we continue to turn back time further, the uncertainties in the initial conditions allow the for the various orbital solutions to diverge from one another. Thus, in one configuration a cluster can impact the stream at a given time and yet at the same moment in a different set of initial conditions, it could be on the other side of the galaxy.

    In total, we report the finding of 73 gaps in our 50 simulations which is on average 1.5 gaps/simulation and were created by 18 different perturbers. The distribution of simulations with a given number of gaps is provided in Table~\ref{table:gap_distribution}.  If we consider NGC~2808 to be an outlier and exclude it from the experiment, we observe an average of 0.6 gaps per simulation. Including all observed gaps results in an average of 1.5 gaps per simulation.

@@

    We need a more sophisticated satistic that can be compared to other simulations and turn to the gap creation rate developed by \citet{2012ApJ...748...20C}. This is the number of gaps per time and normalized by the length of the stream. For our simulations, this rate is given by: \begin{equation} \label{eq:gap_creation_rate} \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the Dirac delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}

    This computation allows us to analyze the distribution of gap creation rates across all simulations. Notice that since the gaps \textit{add in parallel}, naturally gaps that occur at earlier times when the stream was shorter are weighted higher than those that occur when the stream is longer. The results are shown in Fig.~\ref{fig:gapcreationrate}. This is roughly consistent with a simple estimate of the average gap creation rate: with 73 gaps over 5 Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.015~km~s$^{-1}$~kpc$^{-2} = 0.015~\rm{Gyr^{-1}kpc^{-1}}$. This is about double the weighted mean gap creation rate of 0.009~km~s$^{-1}$~kpc$^{-2}$. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}.

@@

    Lastly, we note that of the 73 observed gaps only 8 are in the trailing tail and the rest are in the leading tail. This is a surprising result. Apriori, since the number of star-particles escape at similar rates from the L1 and L2 Lagrange points, each tail is of similar length and density. The main difference between the two tails is that the leading tail is closer to the galactic center than the trailing tail by about 400~pc. Since the lengths are equal, and the offset between the tails is small compared to the galactocentric distance of about 10 kpc, we expected the number of gaps in each tail to be more or less the same. We can compute the the probability of observing the unequal occurances though the binomail distirbution. First, the since the 44 gaps linked to NGC~2808 are the same fly-by, they are not independent events. We remove them from this consideration. This leaves 21 gaps in the leading tail and 8 in the trailing. The probability of observing up to 8 successes in 29 trials, given a 50\% chance of success, is 1.2\%--—unlikely, but possible. Additionally, other perturbers impact at consistent times, which may also violate the assumption of independent events, as seen with NGC~2808.
